---
title: "Advanced Bioinformatics 2026 assessment"
author: "K25163716"
date: "2026-04-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task 3.1

The `:` operator is used to generate a sequence of all integers from 5 to 55, then pass it to the `sum()` function to calculate their total.
```{r}
sum(5:55)
```

## Task 3.2

The function called `sumfun` takes one input `n` and calculates the sum of all integers between 5 and `n` using the `:` operator and `sum()` function.

```{r}
sumfun <- function(n) {
  sum(5:n)
}
```

The function is then repeated for n = 10, n = 20 and n = 100:
```{r}
sumfun(10)
```
```{r}
sumfun(20)
```
```{r}
sumfun(100)
```

## Task 3.3

The Fibonacci sequence starts with 1,1 each subsequent number is the sum of the two preceding numbers (1,1,2,3,5...). The `for` loop is used to calculate and print the first 12 entries.
```{r}
fib <- c(1, 1)
for (i in 3:12) {
  fib[i] <- fib[i-1] + fib[i-2]
}
print(fib)
```

## Task 3.4

Using the `mtcars` dataset a boxplot of miles per gallon (`mgp`) is generated as a function of number of gears (`gear`).To make it readable and colourful `fill` will colour each box by the number of gears. The `factor()` function is used to tell R to treat `gear` as a category rather than a number.
```{r}
library(ggplot2)

ggplot(mtcars, aes(x = factor(gear), y = mpg, fill = factor(gear))) +
  geom_boxplot() +
  labs(x = "Number of Gears", y = "Miles per Gallon", fill = "Gears")
```

## Task 3.5

The `lm()` function is used to fit a linear relationship between speed and braking distance in the `cars` dataset. Speed is measured in **mph** and distance in **feet**.
```{r}
model <- lm(dist ~ speed, data = cars)
summary(model)
```

The fitted intercept is -17.5791 (standard error: 6.7584) and the 
slope is 3.9324 (standard error: 0.4155). This means for every 1 mph 
increase in speed, braking distance increases by approximately 3.93 
feet.

## Task 3.6

The installed `ggplot2` package is used to plot the data points from the `cars` dataset and overlay the linear fit from **Task 3.5** using `geom_smooth()`.
```{r}
ggplot(cars, aes(x = speed, y = dist)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, colour = "blue") +
  labs(x = "Speed (mph)", y = "Braking Distance (feet)")
```

## Task 3.7

To estimate the reaction time, braking distance is assumed as proportional to speed squared. The total stopping distance is modelled as: distance = reaction_time × speed + k × speed². I fit this using `lm()` with no intercept.
```{r}
model2 <- lm(dist ~ 0 + speed + I(speed^2), data = cars)
summary(model2)
```
The coefficient for `speed` is 1.23903, which represents reaction 
distance per mph. Converting to seconds (dividing by 1.467, since 
1 mph = 1.467 ft/s) gives an estimated reaction time of approximately 
0.845 seconds. This is a reasonable result as typical human reaction 
time is between 0.7 and 1.5 seconds. The coefficient for `speed^2` 
is 0.09014, representing the braking component of stopping distance.

The next step involved ploting the data points and the fitted relationship:
```{r}
ggplot(cars, aes(x = speed, y = dist)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ 0 + x + I(x^2), 
              se = FALSE, colour = "red") +
  labs(x = "Speed (mph)", y = "Braking Distance (feet)")
```

## Task 3.8

The count data and sample description files using `read.csv()` and `read.table()` respectively are used.
```{r}
counts <- read.csv("/Users/sylvia/Downloads/LMS_RNAseq_short-master-2023-final 2/course/exercises/data/exercise1_counts.csv", row.names = 1)

sample_info <- read.table("/Users/sylvia/Downloads/LMS_RNAseq_short-master-2023-final 2/course/exercises/data/exercise1_sample_description.info", header = TRUE, sep = "\t")
```

## Task 3.9

`col_data` is created from the sample description file, which will be used to describe the columns of the count matrix. The dimensions of both objects are then checked to make sure they match.
```{r}
col_data <- data.frame(sample_info)
dim(counts)
dim(col_data)
```

The counts matrix contains 26,301 genes across 9 samples. The 
col_data object contains 9 rows (one per sample) and 4 columns 
describing each sample. The number of samples match confirming 
the data is compatible for DESeq2 analysis.

## Task 3.10

A DESeqDataSet object is constructed using the count data and sample description. The `condition` column is used as the design variable, which tells DESeq2 which groups to compare.
```{r}
library(DESeq2)

dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = col_data,
                              design = ~ condition)
dds
```

## Task 3.11

rlog and VST (Variance Stabilizing Transformation) is performed on the DESeq2 object. Both methods normalise the count data to remove the dependence of variance on the mean, making it suitable for visualisation and clustering.
```{r}
rlog_data <- rlog(dds, blind = TRUE)
vst_data <- vst(dds, blind = TRUE)
```

## Task 3.12

A heatmap of the top 40 most highly expressed genes is drawn using 
both rlog and VST transformed data. The `rowMeans()` function is used 
to calculate the average expression per gene and select the top 40. 
Clustering is applied to both rows and columns to reveal patterns in 
the data.
```{r}
library(pheatmap)

# Heatmap using rlog data
top40_rlog <- head(order(rowMeans(assay(rlog_data)), decreasing = TRUE), 40)
pheatmap(assay(rlog_data)[top40_rlog, ], 
         cluster_rows = TRUE, 
         cluster_cols = TRUE,
         main = "Top 40 genes - rlog")
```
```{r}
# Heatmap using VST data
top40_vst <- head(order(rowMeans(assay(vst_data)), decreasing = TRUE), 40)
pheatmap(assay(vst_data)[top40_vst, ], 
         cluster_rows = TRUE, 
         cluster_cols = TRUE,
         main = "Top 40 genes - VST")
```

## Task 3.13

A Sample Distance Matrix (SDM) is generated to visualise the clustering and similarity between samples. The `dist()` function calculates the distances between samples using the rlog transformed data.
```{r}
library(RColorBrewer)

sampleDists <- dist(t(assay(rlog_data)))
sampleDistMatrix <- as.matrix(sampleDists)

colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)

pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors,
         main = "Sample Distance Matrix")
```

The SDM shows clear separation between the mutant and control 
samples. The two groups cluster separately, indicating distinct 
gene expression profiles between conditions. Within each group, 
samples show high similarity to each other as expected.

## Task 3.14

Principal Component Analysis (PCA) is performed using the rlog transformed data. PCA reduces the dimensionality of the data and allows visualisation of the major sources of variation between samples. The percentage of variance explained by each principal component is also reported.
```{r}
plotPCA(rlog_data, intgroup = "condition")
```
```{r}
pca_data <- prcomp(t(assay(rlog_data)))
pca_var <- pca_data$sdev^2 / sum(pca_data$sdev^2) * 100
cat("PC1:", round(pca_var[1], 1), "%\n")
cat("PC2:", round(pca_var[2], 1), "%\n")
```

The PCA plot shows clear separation between the control (FFa) and 
mutant (KOa and KOb) samples along PC1, which explains 70% of the 
variance. PC2 explains 13% of the variance and separates the two 
mutant groups from each other. This confirms the condition is the 
main source of variation in the data.

## Task 3.15

The PCA is repeated using VST transformed data and compared to the rlog PCA from Task 3.14.
```{r}
plotPCA(vst_data, intgroup = "condition")
```