---
title: "Advanced Bioinformatics 2026 assessment"
author: "K25163716"
date: "2026-04-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task 3.1

The `:` operator is used to generate a sequence of all integers from 5 to 55, then pass it to the `sum()` function to calculate their total.
```{r}
sum(5:55)
```

## Task 3.2

The function called `sumfun` takes one input `n` and calculates the sum of all integers between 5 and `n` using the `:` operator and `sum()` function.

```{r}
sumfun <- function(n) {
  sum(5:n)
}
```

The function is then repeated for n = 10, n = 20 and n = 100:
```{r}
sumfun(10)
```
```{r}
sumfun(20)
```
```{r}
sumfun(100)
```

## Task 3.3

The Fibonacci sequence starts with 1,1 each subsequent number is the sum of the two preceding numbers (1,1,2,3,5...). The `for` loop is used to calculate and print the first 12 entries.
```{r}
fib <- c(1, 1)
for (i in 3:12) {
  fib[i] <- fib[i-1] + fib[i-2]
}
print(fib)
```

## Task 3.4

Using the `mtcars` dataset a boxplot of miles per gallon (`mgp`) is generated as a function of number of gears (`gear`).To make it readable and colourful `fill` will colour each box by the number of gears. The `factor()` function is used to tell R to treat `gear` as a category rather than a number.
```{r}
library(ggplot2)

ggplot(mtcars, aes(x = factor(gear), y = mpg, fill = factor(gear))) +
  geom_boxplot() +
  labs(x = "Number of Gears", y = "Miles per Gallon", fill = "Gears")
```

## Task 3.5

The `lm()` function is used to fit a linear relationship between speed and braking distance in the `cars` dataset. Speed is measured in **mph** and distance in **feet**.
```{r}
model <- lm(dist ~ speed, data = cars)
summary(model)
```

The fitted intercept is -17.5791 (standard error: 6.7584) and the 
slope is 3.9324 (standard error: 0.4155). This means for every 1 mph 
increase in speed, braking distance increases by approximately 3.93 
feet.

## Task 3.6

The installed `ggplot2` package is used to plot the data points from the `cars` dataset and overlay the linear fit from **Task 3.5** using `geom_smooth()`.
```{r}
ggplot(cars, aes(x = speed, y = dist)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, colour = "blue") +
  labs(x = "Speed (mph)", y = "Braking Distance (feet)")
```

## Task 3.7

To estimate the reaction time, braking distance is assumed as proportional to speed squared. The total stopping distance is modelled as: distance = reaction_time × speed + k × speed². I fit this using `lm()` with no intercept.
```{r}
model2 <- lm(dist ~ 0 + speed + I(speed^2), data = cars)
summary(model2)
```
The coefficient for `speed` is 1.23903, which represents reaction 
distance per mph. Converting to seconds (dividing by 1.467, since 
1 mph = 1.467 ft/s) gives an estimated reaction time of approximately 
0.845 seconds. This is a reasonable result as typical human reaction 
time is between 0.7 and 1.5 seconds. The coefficient for `speed^2` 
is 0.09014, representing the braking component of stopping distance.

The next step involved ploting the data points and the fitted relationship:
```{r}
ggplot(cars, aes(x = speed, y = dist)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ 0 + x + I(x^2), 
              se = FALSE, colour = "red") +
  labs(x = "Speed (mph)", y = "Braking Distance (feet)")
```
