---
title: "Advanced Bioinformatics 2026 assessment"
author: "K25163716"
date: "2026-04-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Advanced-Bioinformatics-2026")
```

## Task 3.1

The `:` operator is used to generate a sequence of all integers from 5 to 55, then pass it to the `sum()` function to calculate their total.
```{r}
sum(5:55)
```

## Task 3.2

The function called `sumfun` takes one input `n` and calculates the sum of all integers between 5 and `n` using the `:` operator and `sum()` function.

```{r}
sumfun <- function(n) {
  sum(5:n)
}
```

The function is then repeated for n = 10, n = 20 and n = 100:
```{r}
sumfun(10)
```
```{r}
sumfun(20)
```
```{r}
sumfun(100)
```

## Task 3.3

The Fibonacci sequence starts with 1,1 each subsequent number is the sum of the two preceding numbers (1,1,2,3,5...). The `for` loop is used to calculate and print the first 12 entries.
```{r}
fib <- c(1, 1)
for (i in 3:12) {
  fib[i] <- fib[i-1] + fib[i-2]
}
print(fib)
```

## Task 3.4

Using the `mtcars` dataset a boxplot of miles per gallon (`mgp`) is generated as a function of number of gears (`gear`).To make it readable and colourful `fill` will colour each box by the number of gears. The `factor()` function is used to tell R to treat `gear` as a category rather than a number.
```{r}
library(ggplot2)

ggplot(mtcars, aes(x = factor(gear), y = mpg, fill = factor(gear))) +
  geom_boxplot() +
  labs(x = "Number of Gears", y = "Miles per Gallon", fill = "Gears")
```

## Task 3.5

The `lm()` function is used to fit a linear relationship between speed and braking distance in the `cars` dataset. Speed is measured in **mph** and distance in **feet**.
```{r}
model <- lm(dist ~ speed, data = cars)
summary(model)
```

The fitted intercept is -17.5791 (standard error: 6.7584) and the 
slope is 3.9324 (standard error: 0.4155). This means for every 1 mph 
increase in speed, braking distance increases by approximately 3.93 
feet.

## Task 3.6

The installed `ggplot2` package is used to plot the data points from the `cars` dataset and overlay the linear fit from **Task 3.5** using `geom_smooth()`.
```{r}
ggplot(cars, aes(x = speed, y = dist)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, colour = "blue") +
  labs(x = "Speed (mph)", y = "Braking Distance (feet)")
```

## Task 3.7

To estimate the reaction time, braking distance is assumed as proportional to speed squared. The total stopping distance is modelled as: distance = reaction_time × speed + k × speed². This is fitted using `lm()` with no intercept (`~ 0`) because the 
model assumes that at zero speed, stopping distance is zero which 
meaning no intercept is physically meaningful here.
```{r}
model2 <- lm(dist ~ 0 + speed + I(speed^2), data = cars)
summary(model2)
```
The coefficient for `speed` is 1.23903, which represents reaction 
distance per mph. Converting to seconds (dividing by 1.467, since 
1 mph = 1.467 ft/s) gives an estimated reaction time of approximately 
0.845 seconds. This is a reasonable result as typical human reaction 
time is between 0.7 and 1.5 seconds. The coefficient for `speed^2` 
is 0.09014, representing the braking component of stopping distance.

The next step involved ploting the data points and the fitted relationship:
```{r}
ggplot(cars, aes(x = speed, y = dist)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ 0 + x + I(x^2), 
              se = FALSE, colour = "red") +
  labs(x = "Speed (mph)", y = "Braking Distance (feet)")
```

## Task 3.8

The count data and sample description files using `read.csv()` and `read.table()` respectively are used.
```{r}
counts <- read.csv("data/exercise1_counts.csv", row.names = 1)

sample_info <- read.table("data/exercise1_sample_description.info", 
                          header = TRUE, sep = "\t")
```

## Task 3.9

`col_data` is created from the sample description file, which will be used to describe the columns of the count matrix. The dimensions of both objects are then checked to make sure they match.
```{r}
col_data <- data.frame(sample_info)
dim(counts)
dim(col_data)
```

The counts matrix contains 26,301 genes across 9 samples. The 
col_data object contains 9 rows (one per sample) and 4 columns 
describing each sample. The number of samples match confirming 
the data is compatible for DESeq2 analysis.

## Task 3.10

A DESeqDataSet object is constructed using the count data and sample description. The `condition` column is used as the design variable, which tells DESeq2 which groups to compare.
```{r}
library(DESeq2)

dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = col_data,
                              design = ~ condition)
dds
```

## Task 3.11

rlog (regularised log) and VST (Variance Stabilizing Transformation) 
transformations are performed on the DESeq2 object. Raw count data 
from RNA-seq experiments has a mean-variance relationship where highly 
expressed genes show much higher variance than lowly expressed ones. 
Both transformations address this by stabilising the variance across 
the range of mean expression values, making the data suitable for 
visualisation and clustering. rlog is preferred for smaller datasets 
(n < 30 samples) as it is more robust, while VST is faster and better 
suited to larger datasets. Both are applied here for comparison.
```{r}
rlog_data <- rlog(dds, blind = TRUE)
vst_data <- vst(dds, blind = TRUE)
```

## Task 3.12

A heatmap of the top 40 most highly expressed genes is drawn using 
both rlog and VST transformed data. The top 40 genes are selected 
using `rowMeans()` to calculate average expression per gene focusing 
on the most highly expressed genes allows clearer visualisation of 
expression patterns across samples. Hierarchical clustering is applied 
to both rows (genes) and columns (samples) to group genes with similar 
expression profiles and reveal whether samples cluster by condition. 
This is a useful quality control step to check that biological 
replicates behave consistently and that conditions are separating 
as expected.
```{r}
library(pheatmap)

# Heatmap using rlog data
top40_rlog <- head(order(rowMeans(assay(rlog_data)), decreasing = TRUE), 40)
pheatmap(assay(rlog_data)[top40_rlog, ], 
         cluster_rows = TRUE, 
         cluster_cols = TRUE,
         main = "Top 40 genes - rlog")
```
```{r}
# Heatmap using VST data
top40_vst <- head(order(rowMeans(assay(vst_data)), decreasing = TRUE), 40)
pheatmap(assay(vst_data)[top40_vst, ], 
         cluster_rows = TRUE, 
         cluster_cols = TRUE,
         main = "Top 40 genes - VST")
```

## Task 3.13

A Sample Distance Matrix (SDM) is generated to visualise the 
similarity and dissimilarity between all pairs of samples. The 
`dist()` function calculates Euclidean distances between samples 
using the rlog transformed data, the smaller the distance, the 
more similar two samples are. This is an important quality control 
step as it allows us to verify that biological replicates cluster 
together and that the two conditions (control and mutant) are 
clearly separated from each other. The `RColorBrewer` package is 
used to apply a blue colour scale where darker blue indicates 
greater similarity (smaller distance) and lighter blue indicates 
greater dissimilarity.
```{r}
library(RColorBrewer)

sampleDists <- dist(t(assay(rlog_data)))
sampleDistMatrix <- as.matrix(sampleDists)

colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)

pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors,
         main = "Sample Distance Matrix")
```

The SDM confirms a clear separation between the mutant and control 
samples. The two groups cluster separately, indicating distinct 
gene expression profiles between conditions. Within each group, 
samples show high similarity to each other as expected for 
biological replicates. This gives confidence that the data is 
of good quality and suitable for downstream differential 
expression analysis.

## Task 3.14

Principal Component Analysis (PCA) is performed using the rlog 
transformed data. PCA is a dimensionality reduction technique that 
transforms the data into a set of uncorrelated variables called 
principal components, ordered by the amount of variance they 
explain. This allows visualisation of the major sources of variation 
between samples in a 2D plot. PC1 captures the largest source of 
variation in the data, and PC2 the second largest. If the experiment 
is well designed, samples should cluster by condition along PC1. 
The `plotPCA()` function from DESeq2 uses the top 500 most variable 
genes by default, which focuses the analysis on the most 
informative genes.
```{r}
plotPCA(rlog_data, intgroup = "condition")
```
```{r}
pca_data <- prcomp(t(assay(rlog_data)))
pca_var <- pca_data$sdev^2 / sum(pca_data$sdev^2) * 100
cat("PC1:", round(pca_var[1], 1), "%\n")
cat("PC2:", round(pca_var[2], 1), "%\n")
```

The PCA plot shows clear separation between the control (FFa) and 
mutant (KOa and KOb) samples along PC1, which explains 70% of the 
variance — confirming that the experimental condition is the dominant 
source of variation in the dataset. PC2 explains 13% of the variance 
and separates the two mutant subgroups (KOa and KOb) from each other, 
suggesting a secondary source of biological variation within the 
mutant condition. Together PC1 and PC2 explain 83% of the total 
variance, indicating a well-structured dataset.

## Task 3.15

The PCA is repeated using VST transformed data to allow comparison 
with the rlog method used in Task 3.14. While both transformations 
aim to stabilise variance, they use different mathematical approaches 
and may produce slightly different results. Comparing the two PCA 
plots allows assessment of whether the choice of normalisation method 
affects the biological conclusions drawn from the data.
```{r}
plotPCA(vst_data, intgroup = "condition")
```

The VST PCA shows PC1 explaining 69% of variance and PC2 explaining 
14% of variance — almost identical to the rlog results (PC1: 70%, 
PC2: 13%). Both plots show the same overall pattern of separation 
between control and mutant samples along PC1, with PC2 separating 
the two mutant subgroups. This consistency between methods confirms 
that the biological conclusions are robust and not dependent on the 
choice of normalisation method. VST is generally preferred for larger 
datasets due to its computational efficiency, while rlog is more 
accurate for smaller sample sizes such as this one.